### 2020-02-16 doc/icfp20/counterpoint.tex
figures
move figures to fig/

\newcommand{\FS}{
  \begin{figure}
    \includegraphics[width=11cm]{fs.png}
    \caption{First Species Counterpoint}
    \label{fig:fs}
  \end{figure}
}

\renewcommand{\SS}{
  \begin{figure}
    \includegraphics[width=10cm]{ss.png}
    \caption{Second Species Counterpoint}
    \label{fig:ss}
  \end{figure}
}

@@

\FS

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which is required to start with a tonic and consist only of whole notes.
Figure~\ref{fig:fs} shows an example of first species counterpoint.
The lower line is the cantus firmus, which we excerpt from a famous
German song called ``Froschgesang'' (Frog's song).
The upper line is the counterpoint composed by the second author.

@@

The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.
As shown in Figure~\ref{fig:motion}, there are four kinds of motion,
and among them, parallel and similar motion makes it harder to
distinguish between two voices, especially when the second interval
is a perfect one.
To avoid this problematic motion, we define the \texttt{checkMotion}
function, which lists all invalid motion approaching a perfect interval.

@@

\SS

We next discuss a more complex variant of counterpoint, called the
second species.
In second species, we set \emph{two} half notes against every cantus
firmus note.
Figure~\ref{fig:ss} shows an example of such two-against-one
counterpoint, again composed for the Frog's song.
As can be seen from this example, the first and last bars may have
a different structure from middle bars.
More specifically, it is encouraged to begin the counterpoint line with
a half rest and end with a whole note.
Therefore, in our implementation, we reuse the \texttt{PitchInterval} 
type for the first and last bars, and define a new type
\texttt{PitchInterval2} for middle bars:

>>>>

\newcommand{\FS}{
  \begin{figure}
    \includegraphics[width=6cm]{fig/fs.png}
    \caption{First Species Counterpoint}
    \label{fig:fs}
  \end{figure}
}

\newcommand{\Motion}{
  \begin{figure}
    \includegraphics[width=6cm]{fig/motion.png}
    \hspace{1cm}
    \includegraphics[width=3cm]{fig/badmotion.png} \\
    %\vspace{-1cm}
    \begin{flushleft}
    \begin{small}
      \hspace{1.95cm} contrary \ \ \ \ parallel \ \ \ \ \ \ \ similar \ \ \ \ \ \ oblique
      \hspace{2.2cm} 5th \hspace{9.5mm} 8th
    \end{small}
    \end{flushleft}
    \caption{Left: Four Kinds of Motion, Right: Invalid Motion}
    \label{fig:motion}
  \end{figure}
}

\renewcommand{\SS}{
  \begin{figure}
    \includegraphics[width=7cm]{fig/ss.png}
    \caption{Second Species Counterpoint}
    \label{fig:ss}
  \end{figure}
}

@@

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which is required to start with a tonic and consist only of whole notes.
Figure~\ref{fig:fs} shows an example of first species counterpoint:

\begin{figure}[h]
  \includegraphics[width=6cm]{fig/fs.png}
  \caption{First Species Counterpoint}
  \label{fig:fs}
\end{figure}

\noindent The lower line is the cantus firmus, which we excerpt from
a famous German song called ``Froschgesang'' (Frog's song).
The upper line is the counterpoint composed by the second author.

@@

The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.

\begin{figure}
    \includegraphics[width=6cm]{fig/motion.png}
    \hspace{1cm}
    \includegraphics[width=3cm]{fig/badmotion.png} \\
    \begin{flushleft}
    \begin{small}
      \hspace{1.95cm} contrary \ \ \ \ parallel \ \ \ \ \ \ \ similar \ \ \ \ \ \ oblique
      \hspace{2.2cm} 5th \hspace{9.5mm} 8th
    \end{small}
    \end{flushleft}
    \caption{Four Kinds of Motion (left) and Bad Motion (right)}
    \label{fig:motion}
  \end{figure}

As can be seen from the figure on the left, there are four kinds of
motion: \emph{contrary} (two lines go in different directions),
\emph{parallel} (two lines go in the same direction by the same
distance), \emph{similar} (two lines go in the same direction),
and \emph{oblique} (one line plays the same note).
Among these variants, parallel and similar motion makes it harder
to distinguish between two voices, especially when the second
interval is a perfect one, as in the figure on the right.
To avoid problematic motion, we define the \texttt{checkMotion}
function, which lists all invalid motion approaching a perfect interval.

@@

We next discuss a more complex variant of counterpoint, called the
second species.
In second species, we set \emph{two} half notes against every cantus
firmus note.
Figure~\ref{fig:ss} is an example of such two-against-one counterpoint,
again composed for the Frog's song:

\begin{figure}[h]
  \includegraphics[width=7cm]{fig/ss.png}
  \caption{Second Species Counterpoint}
  \label{fig:ss}
\end{figure}

As we can see from this example, the first and last bars may have
a different structure from middle bars.
More specifically, it is encouraged to begin the counterpoint line with
a half rest and end with a whole note.
Therefore, in our implementation, we reuse the \texttt{PitchInterval} 
type for the first and last bars, and define a new type
\texttt{PitchInterval2} for middle bars:
